---
title: "Sleep Staging: Live Walkthrough (SC4001E0)"
description: |
  End-to-end automatic AASM sleep staging on a real 22-hour Sleep-EDF
  cassette recording, with posterior probability figures.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 10,
  fig.height = 4,
  dpi       = 150,
  out.width = "100%"
)
```

```{r packages, message = FALSE}
library(mrpheus)
library(dplyr)
library(tidyr)
library(ggplot2)
```

This article runs the complete mrpheus sleep staging pipeline on
**SC4001E0-PSG.edf** from the Sleep-EDF Cassette study (Kemp et al., 2000) —
a 22-hour ambulatory PSG of a healthy young adult.
For download instructions see `vignette("sleep-staging")`.

The Sleep-EDF Cassette recordings start mid-afternoon and capture a full
sleep-wake cycle plus the following morning. Around 75% of the recording is
daytime Wake; the sleep period is approximately 7–8 hours.

---

## Pipeline

The code below shows the full pipeline. Results are loaded from a
pre-computed dataset bundled with the package (the EDF is too large to
include, but the staging output is identical).

```{r pipeline, eval = FALSE}
rec <- read_edf("~/mne_data/physionet-sleep-data/SC4001E0-PSG.edf")
psg <- prepare_psg(rec)

staging <- stage_epochs(
  psg,
  eeg_channel = "EEG Fpz-Cz",
  eog_channel = "EOG horizontal",
  emg_channel = "EMG submental"
)
```

```{r load-precomputed, include = FALSE}
staging <- readRDS(
  system.file("extdata", "SC4001E0_staging.rds", package = "mrpheus")
)
```

```{r peek}
staging
```

---

## Stage distribution

```{r palette, include = FALSE}
stage_pal <- c(
  W   = "#AAAAAA",   # neutral grey — awake
  N1  = "#B07C3A",   # ochre — transition
  N2  = "#6A7840",   # olive — light sleep
  N3  = "#014370",   # navy — deep sleep
  REM = "#B83E2C"    # vermillion — active dreaming
)
```

```{r distribution-table}
stage_summary <- staging |>
  count(stage) |>
  mutate(
    stage      = factor(stage, levels = c("W", "N1", "N2", "N3", "REM")),
    duration_h = round(n * 30 / 3600, 2),
    pct        = round(n / nrow(staging) * 100, 1)
  ) |>
  arrange(stage)

stage_summary
```

```{r fig-distribution, fig.height = 3.5}
stage_summary |>
  ggplot(aes(x = stage, y = duration_h, fill = stage)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0(duration_h, " h\n(", pct, "%)")),
            vjust = -0.4, size = 3.2, lineheight = 1.1) +
  scale_fill_manual(values = stage_pal) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    x     = NULL,
    y     = "Duration (hours)",
    title = "Sleep Stage Distribution — SC4001E0",
    subtitle = sprintf(
      "%.1f-hour cassette recording  ·  %d × 30-second epochs",
      nrow(staging) * 30 / 3600, nrow(staging)
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "#7C5432", size = 10)
  )
```

Wake accounts for the majority of the recording because the cassette started
at ~16:13 local time and ran continuously through the night and into the
following afternoon.

---

## Posterior probability heatmap

Each epoch carries five posterior probabilities (one per AASM stage) that sum
to 1. Visualising all five simultaneously over the full recording reveals both
the overall structure of the night and the model's confidence at each point
in time.

```{r fig-heatmap-full, fig.height = 3.5}
staging |>
  select(epoch, starts_with("prob_")) |>
  pivot_longer(starts_with("prob_"), names_to = "stage", values_to = "prob") |>
  mutate(
    stage  = factor(sub("prob_", "", stage), levels = c("N3", "N2", "N1", "REM", "W")),
    time_h = (epoch - 1L) * 30 / 3600
  ) |>
  ggplot(aes(x = time_h, y = stage, fill = prob)) +
  geom_tile() +
  scale_fill_gradient(
    low  = "#F7F3EB",
    high = "#3C2212",
    name = "Posterior\nprobability",
    guide = guide_colourbar(barheight = 4)
  ) +
  labs(
    x        = "Time from recording start (hours)",
    y        = NULL,
    title    = "Sleep Staging Posterior Probabilities",
    subtitle = "SC4001E0 — full 22-hour recording"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.subtitle = element_text(colour = "#7C5432", size = 10),
    panel.grid    = element_blank(),
    axis.text.y   = element_text(face = "bold")
  )
```

The dark band in the **W** row spanning the first ~8 hours and last ~5 hours
reflects the daytime Wake periods before and after sleep. The sleep period
(roughly hours 8.5–17) is visible as sustained activity in the **N2**, **N3**,
and **REM** rows.

---

## Sleep period in detail

Zooming into the sleep window (starting from just before lights-out at
~30,600 s ≈ hour 8.5) makes the ultradian architecture clearer.

```{r fig-heatmap-sleep, fig.height = 3.5}
staging |>
  filter(epoch >= 1001L, epoch <= 1950L) |>
  select(epoch, starts_with("prob_")) |>
  pivot_longer(starts_with("prob_"), names_to = "stage", values_to = "prob") |>
  mutate(
    stage  = factor(sub("prob_", "", stage), levels = c("N3", "N2", "N1", "REM", "W")),
    time_h = (epoch - 1L) * 30 / 3600
  ) |>
  ggplot(aes(x = time_h, y = stage, fill = prob)) +
  geom_tile() +
  scale_fill_gradient(
    low  = "#F7F3EB",
    high = "#3C2212",
    name = "Posterior\nprobability",
    guide = guide_colourbar(barheight = 4)
  ) +
  labs(
    x        = "Time from recording start (hours)",
    y        = NULL,
    title    = "Sleep Period — Posterior Probabilities",
    subtitle = "SC4001E0 · lights-out to end of sleep"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.subtitle = element_text(colour = "#7C5432", size = 10),
    panel.grid    = element_blank(),
    axis.text.y   = element_text(face = "bold")
  )
```

---

## Model confidence

High posterior probabilities indicate epochs the model is unambiguous about.
Lower values reflect genuine signal ambiguity at stage transitions or in
epochs with mixed features.

```{r fig-confidence, fig.height = 3.5}
staging |>
  mutate(
    stage      = factor(stage, levels = c("W", "N1", "N2", "N3", "REM")),
    confidence = pmax(prob_N1, prob_N2, prob_N3, prob_REM, prob_W)
  ) |>
  ggplot(aes(x = confidence, fill = stage)) +
  geom_histogram(bins = 60, colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = stage_pal) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    x        = "Max posterior probability",
    y        = "Number of 30-second epochs",
    fill     = "Predicted stage",
    title    = "Model Confidence Distribution",
    subtitle = "SC4001E0 — 2,650 epochs"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(colour = "#7C5432", size = 10))
```

Most epochs concentrate near 1.0, reflecting high model confidence. The
lower-confidence tail consists primarily of **N1** epochs and brief
transitions between stages.

---

## Passing to hypnor

The `staging` tibble can be passed directly to **hypnor** for hypnogram
visualisation and sleep architecture metrics:

```{r export, eval = FALSE}
hypnogram <- export_hypnogram(
  staging,
  epoch_s        = 30,
  start_time     = rec$header$startTime,
  participant_id = "SC4001"
)

# hypnor::plot_hypnogram(hypnogram)
# hypnor::compute_architecture(hypnogram)
```

---

## References

Kemp, B., Värri, A., Rosa, A. C., Nielsen, K. D., & Gade, J. (2000).
A simple format for exchange of digitized polygraphic recordings.
*Electroencephalography and Clinical Neurophysiology*, 69(5), 391–397.
<https://doi.org/10.1016/0013-4694(87)90048-7>

Vallat, R., & Walker, M. P. (2021). An open-source, high-performance tool
for automated sleep staging. *eLife*, 10, e70092.
<https://doi.org/10.7554/eLife.70092>
